Temporal ordering and registration of images in studies of developmental dynamics

### 2015-01-31 image_analysis_paper_development2/manuscript.tex
updated paper

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology. 
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis; here, the correct rotational orientation and temporal order are independently known and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent. 
%
Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can extend to three-dimensional imaging. 
%
For data sets from multiple samples, we will also show how to compute an average trajectory from a set of registered and ordered images to obtain a smooth description of the underlying developmental dynamics.

>>>>

We present a robust algorithmic approach to registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression for temporal ordering.
%
Our approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing embryos in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology. 
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent. 
%
Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods extend to three-dimensional imaging. 
%
We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.
